Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals

### 2022-10-08 draft_BLC/BLC_AM-ID_manuscript.tex
shortening

Note that ``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested. Recent research suggests that (see \cite{sirsa2013effects} for a review). IE has a target phonology that is distinct from the varied L1s of its speakers, as well as from other native varieties of English. For the present study, we focus on ``General(ized) Indian English" \citep{masica1972sound}. This sound system is described in section \ref{vowel systems}.

@@

\begin{enumerate}[label=(\roman*)]
	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
	\item Given the rarity of IE phonology without L1 `influence' (it is largely spoken as an L2) it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a baseline condition.
\end{enumerate}

The socio-linguistic facts about English in India also suggest that, make mixed-language processing of English and an L1 an ecologically valid paradigm to understand short-term phonetic interaction.

\subsubsection*{Vowel systems} \label{vowel systems}
The vowel inventory of Western Bangla (the variety spoken by the participants in this study) consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}. The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1$\times$F2 formant space. Note that there is no mid-central vowel category.

@@

Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[c.f.][]{grosjean1998studying}. Thus, we elicited all utterances in a bilingual language mode. Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.

@@

Research comparing bilingual speakers' phonological systems with monolinguals treats them as relatively stable over time. The difference from monolingual norms is interpreted as the cumulative result of cross-language influence over long periods \citep[e.g.][]{guion2003vowel, caramazza1973acquisition}. However, longitudinal studies %(citations) 
and between-subject comparisons of bilingual speakers with different durations of L2 exposure suggest that interaction between the sound systems is dynamic: over time, increasing exposure to an L2 can reduce effects of L1 influence, leading to a more fine-grained separation and native-like production of foreign contrasts. For example, \cite[][vowel quality]{bohn1992production} 
observed that approximately 7 years' difference in L2 exposure between groups having the same age of acquisition led to a significant difference in the amount of cross-language influence. These have been interpreted as changes to long-term phonological representations.

Moreover, changes due to transfer are not necessarily unidirectional or irreversible. \cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language. This not only evidenced that cross-language interaction can be triggered in an order of months, but also that the effects can be reversed within a similar time range. A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks). Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2. For some sounds and features (though not others, cf. sec.\ref{asymmetry between sounds}), transfer was additive over time. How long these effects last in the absence of regular L2 input was not tested. 
% if effects are cumulative, then the difference between switched and non-switched tokes should be lesser in later test blocks (previous two paragraphs needed?)

@@

\subsubsection*{L1 vs. L2 status} 
Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} posits that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another. Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories. In non-switched production, this is evidenced through a pervasive L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf. \textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.

@@

\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups: native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition. This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer. They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}. Since L1 is the `stronger' language, the amount of inhibition required on L1 while using an L2 is greater than the inhibition required on an L2 while using the L1. This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2. Lower switch cost results in a lack of visible cross-language transfer effects on L2. \cite{tsui2019impact} report comparable results, but only in participants who were not equally dominant in both languages. Balanced bilinguals did not demonstrate any transfer effects in VOT. They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.

@@

In the present study, our research question concerns whether L2 vowels are produced differently in nonswitch vs switch conditions. Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted. Although our primary research question merely concerns whether productions in the two conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}. The procedure is described below:

\subsubsection*{Baseline Bengali data from the SHRUTI corpus}

The SHRUTI corpus is the largest free corpus of Western Bengali speech that we are aware of. It comprises of 21.64 hours of read speech, containing 7383 unique sentences (22012 unique words) from Bengali newspapers in different domains, covering the most frequently used words of the language. The speakers are 34 adults (age range 20-40, 75\% male, 25\% female) who have grown up in the Indian state of West Bengal, and speak standard colloquial Bengali. All of the speakers have university-level education or higher. Given these parallels in linguistic, regional, and socio-economic background, we believe that Bengali speech from this corpus is representative of the average target Bengali vowel categories that are of interest in this study.
To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target. Next, we compared these to our participants' productions in the non-switched English condition to understand the relative positions of these categories. This is visualized in figure \ref{vowels_unilingual_ie_bengali}.

Note that we refrained from using corpus data for average Indian English vowel positions, opting instead to use the participant data from this study, because IE has high regional variability (c.f. discussion in section \ref{bengali_english_in_india}). To the best of our knowledge, there is no freely available corpus of IE speech from Bengali speakers in India.

The F1$\times$F2 plot indicates that the Bengali category \nt{\ae} is more fronted, and \nt{a:} lower and more fronted, compared to the English categories \nt{\ae} and \nt{2} respectively.

@@

Given our inclusion criteria, we were able to recruit 10 participants for this study. Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size. As detailed above, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7\%) in the final study. Note that the criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies. This demonstrates the degree of linguistic variability in the present population. Norms for experimental research on bilingualism have largely been established in a Western context. In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity (c.f. discussion in section \ref{introduction}). A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f. section \ref{data_availability}). We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.

@@

Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs). Thus, Vowel is a fixed effect in the models. The experimental treatment of interest is Context, i.e. whether the English target item was presented in an English (nonswitch) or Bengali (switch/mixed) context. Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance. We hypothesize that the direction of such shift in the mixed condition, if any, is towards a corresponding L1 category. Therefore, we expect the effect of Context on formant frequencies, i.e. the direction of shift, to be moderated by the vowel category (c.f. the discussion in Section \ref{asymmetry between sounds}, leading to our second research question \textbf{Asymmetry}). Thus, the main effect of interest is the interaction term Context*Vowel. This corresponds to the transfer pattern. Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer. If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction. Since we measured FFs at five points in the vowel, Time is expected to affect the FF. However, the precise effect of formant dynamics is expected to depend on the vowel category. Thus, we expect the interaction of Vowel*Time to predict formant frequency. If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.

@@

Thus, the optimal model was: f1 $\sim$ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject). Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model. p-vales smaller than 0.05 are starred.

@@

Independent of experimental condition, there is a significant effect of Task: the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task. This is surprising, as we have no \textit{a priori} reason to expect a uniform difference in vowel quality across tasks. A significant interaction with Time is strongest at the beginning of the vowel, indicatin a local effect. This is consistent with the interpretation that the unexpected Task effect is an artifact of the paradigm. A significant Vowel*Context*Task interaction indicates a difference in transfer patterns such that transfer in the F1 (height) dimension is greater in the code-switching (s) task compared to the picture-naming (p) task.\\

@@

\subsection{Vowel backness -- F2}

Thus, the optimal model is: f2 $\sim$ vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject). Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model. p-values smaller than 0.05 are in starred.

@@

As with F1, we find that Time significantly influences F2 values in the model, reflecting the dynamic nature of speech which causes acoustic variation over the course of articulation. Moreover, the Time*Vowel interaction shows that this trajectory differs across the two vowels: while the positive slope for Time indicates that F2 increases i.e. vowels are fronted with time, the negative slope of the Time*Vowel interaction shows that this effect is smaller for \nt{2}. This is expected, as figure \ref{vowels_unilingual_ie_bengali} confirms that the category \nt{\ae} is overall more fronted than \nt{2}.

Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e. being produced as more backed, than \nt{\ae}. As with vowel height, there is no significant independent effect of Context, showing that mixed language use does not lead to uniform fronting or backing across vowels. This is not surprising, as we expect movement to be towards the corresponding L1 category, and thus differ across vowels. The negative slope of the estimate indicates that on average, vowels have lower F2 i.e. are backed, in the nonswitch (e) condition. In other words, vowels are fronted in the mixed condition. This effect is greater in \nt{\ae}, as indicated by the negative slope of the Vowel*Context interaction. In the F2 dimension, too, the lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time, suggesting that the underlying acoustic targets are different in the two conditions.

We see a significant effect of Task on F2, such that vowels in the code-switching (s) task have lower F2, i.e. are more backed than those in the picture-naming (p) task. As with F1, this is an unexpected result; we have no principled reason to expect code-switching to induce an across-board backing of vowels. Vowels are backed in the code-switching task, independent of experimental condition. A significant interaction with Time, with a positive slope for the estimate, shows that this effect is local: it is strongest at the beginning of the vowel, and reduces over the course of the vowel. This aligns with the possibility that the unexpected Task effect is an artifact of the paradigm and the particular stimuli used. The Vowel*Context*Task interaction is of interest, as it relates to our research question of how the test paradigm affects transfer. Unlike vowel height, we do not find patterns of shift in vowel backness (F2) to differ across paradigms, as indicated by the lack of a significant effect for the fixed effect.

@@

Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition. Based on previous literature, we expect that such shifts in vowel categories result from increased cross-language influence during mixed-language use, i.e. shifts are targeted with respect to L1 categories. As discussed in section \ref{asymmetry between sounds}, although our primary research question is about whether productions in switch vs nonswitch conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. Figure \ref{vowels_unilingual_ie_bengali} shows the non-switched IE productions of the participants in relation to average positions of related vowel categories in Bengali. Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing F1$\times$F2 vowel plots for IE vowel productions in the nonswitch and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}. Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses. However, figure \ref{vowels_e_b} provides a heuristic confirmation for our informal hypotheses in section \ref{asymmetry between sounds}) that the IE \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}. The following sections discuss the implications of these results.

@@

The main aim of this study was to examine short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals. Two interrelated questions, discussed in section \ref{introduction}, bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2. Together, these serve as the basis for our theories about the language systems of multilingual individuals exist in relation to one another, and consequently, what cognitive models of language should look like.

\paragraph{\textbf{What causes transfer: language mode vs switching}}
In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session, such that the language mode (participants' global language ``setting") remained the same throughout. In spite of this, results demonstrate consistent differences between switched and non-switched tokens. This confirms that the local act of switching does trigger transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}. These studies all examined the effect of switching on a temporal feature of speech, namely voice onset time or VOT. The results of the present study indicate that the same happens in vowels. Moreover, the exact effect of transfer is interesting: VOT is a one-dimensional measure. Thus, observing an increase or decrease in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms. Since vowel quality measures show directionality as well as magnitude, we can observe finer details in the patterns of shift. In the present study, we find that productions of English vowels do not merely shift towards a Bengali counterpart, but rather towards a set of Bengali-like \textit{norms}: a vowel category that doesn't exist in Bengali nonetheless moves towards a related (perceptually close) Bengali category. This result parallels \cite{simonet2014phonetic}, who found the Catalan \nt{O} of bilingual speakers shifts towards a related Spanish category, \nt{o}, in a bilingual language mode (note that they did not examine code-switched speech, and focused on vowel height as a target of transfer). The study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2 (moderated by language dominance). This is interpreted as a follows: a bilingual mode temporarily co-activates both language systems, making both L1 and L2 representations available during processing. These representations interact while producing speech in such a language mode, leading to cross-language phonetic transfer. But what does this interaction look like, and how are these abstract representations processed during the act of producing speech? Examining speech during switching can shed light on this.

The time course of such shifts provides an insight into this. Vowels in the mixed condition do not start out the same, eventually deviating in acoustics from their nonswitch counterparts. Instead, we see a difference between the conditions right from the start of articulation. This suggests that such interaction involves the speech planning stage. The pattern of transfer observed in this study, particularly for \nt{2}, also demonstrates that beyond a simple L2-L1 mapping, transfer during online processing there makes reference to entities beyond the level of the individual sound. Specifically, such transfer demonstrates a system-level interaction: a sound that does not exist in the L1 repertoire nonetheless acquires a different target, and the target is related to the L1 system in a meaningful way. This cannot be reasonably explained as articulatory inertia carried over from uttering L1 sounds. It is also difficult to account for in a purely exemplar framework, as the discourse does not contain any instances of ``L1 \nt{2}" that could affect the L2 category. If representations are combined in creative ways during online speech processing within an utterance, then can we think of intermediate representations not just for individual sounds, but rather for sound systems? While more detailed empirical data will eventually enable us to clarify such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, for example as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children. This visualizes the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures. We propose that vowels are a good feature to examine the plausibility of such systems as cognitive models for language processing: in addition to continuous measures (unlike phonological variables), they also provide two-dimensional data (unlike VOT) that is suited to further test and refine predictions from such a model.

@@

\subsection{Asymmetries in Phonetic Transfer}
The literature on short-term phonetic transfer has frequently noted differences in transfer patterns between different sound categories in a language. This has been attributed to both general linguistic principles such as contrast preservation, and language-specific factors such as frequency and distribution of categories. Examining these asymmetries more closely, and understanding their sources, helps us figure out what factors affect/moderate transfer and thus, by extension, what level of representation and processing this interaction happens at, and the nature of short-term intra-speaker variation. 
The present study examined transfer in two L2 categories, \nt{2} and \nt{\ae}. We hypothesized an asymmetrical shift between the two vowels (\textbf{Asymmetry}), owing to differences in:
 \begin{itemize}
 	\item their position in the Indian English vowel space: \nt{\ae} is in a more crowded part of the vowel space
 	\item the presence/absence of a corresponding category in the L1 Bengali inventory: \nt{\ae} is a common phonological category in Bengali and English, while \nt{2} does not exist as a category in Bengali
 \end{itemize}

>>>>

``Indian English" is not a monolithic entity; there are large regional variations in the English speech of Indian speakers, and the literature on IE phonology is thus neither ``standardized" nor uncontested. Recent research suggests that IE has a target phonology that is distinct from the varied L1s of its speakers, as well as from other native varieties of English (see \cite{sirsa2013effects} for a review). For the present study, we focus on ``General(ized) Indian English" \citep{masica1972sound} (\ref{vowel systems}).

@@

\begin{enumerate}[label=(\roman*)]
	\item In multilingual populations, long-term representations of sound categories are expected to be affected by multiple languages. 
	\item Given the rarity of IE phonology without L1 `influence' (it is almost exclusively spoken as an L2) it is more meaningful to think of cross-language transfer in L2 as relative to a speaker's own production in a baseline condition.
\end{enumerate}

The socio-linguistic facts about English in India also suggest that mixed-language processing of English and an L1 an ecologically valid paradigm to understand short-term phonetic interaction.

\subsubsection*{Vowel systems} \label{vowel systems}
The vowel inventory of Western Bangla consists of \nt{i, e, \ae, a, O, o, u}, and their nasalized counterparts \citep{garry2001facts}. The vowel chart in figure \ref{vowels_bengali} shows their distribution in the F1$\times$F2 formant space. Note that there is no mid-central vowel category.

@@

Since participants of this study are in an environment that largely contains mixed-language input and multilingual interlocutors, we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode \citep[c.f.][]{grosjean1998studying}. Thus, we elicited all utterances in the same bilingual test block. Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.

@@

Longitudinal studies and between-subject comparisons of speakers with different durations of L2 exposure suggest that interaction between bilingual sound systems is dynamic \citep[e.g.][]{bohn1992production}. Moreover, changes due to transfer are not necessarily unidirectional or irreversible. \cite[][VOT]{sancier1997gestural} first demonstrated that spending 2-5 months in an L1 or L2 environment causes productions in both languages to ``drift" towards the ambient language, showing that cross-language interaction can be triggered in an order of months, and also reversed within a similar time range. A more recent study by \cite[][VOT]{tobin2017phonetic} reported comparable effects in an even shorter duration (2-4 weeks). Based on a short-term longitudinal study, \cite[][VOT]{chang2012rapid} demonstrates that over the course of the first five weeks of learning an L2, there is a gradual convergence of L1 towards L2. For some sounds and features (though not others, cf. sec.\ref{asymmetry between sounds}), transfer was additive over time. How long these effects last in the absence of regular L2 input was not tested.

@@

\subsubsection*{L1 vs. L2 status} 
Flege's Speech Learning Model (SLM) \citeyearpar{flege1995second,flege2007language} proposes that the sound categories of a bilingual speaker exist in a common phonological space, and therefore in principle, both L1 and L2 categories can influence one another. Thus, the patterns of cross-language transfer depend on how L2 phonemes are mapped in relation to the existing L1 categories. In non-switched production, this is evidenced through a L1 influence on non-native contrasts that are perceptually linked to an existing L1 contrast (cf. \textsc{equivalence classification}: \cite{flege1984limits,flege1987production}), and the observation that both L1 and L2 sound systems of bilinguals differ from those of corresponding monolingual speakers \citep[e.g.][vowel quality]{guion2003vowel}.

@@

\cite{olson2013bilingual} first reported a unidirectional VOT shift of L1 towards L2 in two different groups: native speakers of Spanish (short-lag) and English (long-lag), matched for proficiency and age of L2 acquisition. This established that beyond language-specific differences, the L1 vs L2 status of the language does mediate transfer. They interpret this asymmetry in terms of the Inhibitory Control Model (ICM) \citep{green1998mental}: to select a phonetic realization from one language, the other must be \textsc{inhibited}. Greater inhibition is required on L1 while using an L2 than the reverse. This greater initial inhibition on L1 means that switching into L1 incurs a greater \textsc{switch cost} (more cross-language influence) than switching into L2. Lower switch cost results in a lack of visible cross-language transfer effects on L2. \cite{tsui2019impact} report comparable results, but only in participants who were not equally dominant in both languages. Balanced bilinguals did not demonstrate any transfer effects in VOT. They propose that this is because balanced bilinguals have better inhibitory control (low switch cost in both languages) due to greater experience with language switching.

@@

In the present study, our research question concerns whether L2 vowels are produced differently in nonswitch vs switch conditions. Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted. Although our primary research question merely concerns whether productions in the two conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}. We elaborate on the corpus details and procedure in the Supplementary materials.

We extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, and compared these to our participants' productions in the non-switched English condition. This is visualized in figure \ref{vowels_unilingual_ie_bengali}. 
The F1$\times$F2 plot indicates that the Bengali category \nt{\ae} is more fronted, and \nt{a:} lower and more fronted, compared to the English categories \nt{\ae} and \nt{2} respectively.

@@

Given our inclusion criteria, we were able to recruit 10 participants for this study. Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size. As detailed above, even after a recruitment call that already controlled for language background and education level, we were able to include only 10 out of 28 volunteers (35.7\%) in the final study. The criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies. This demonstrates the degree of linguistic variability in the present population. Norms for experimental research on bilingualism have largely been established in a Western context. In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity. A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f. section \ref{data_availability}). We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.

@@

Our study involved two IE vowels. Vowel is a fixed effect, since it is expected to systematically affect formant frequencies (FFs). The experimental treatment is Context, i.e. whether the English target item was presented in an English (nonswitch) or Bengali (switch/mixed) context. Our first research question (\bf{RQ1---L2 vowels}) asks whether L2 vowel quality of proficient bilinguals is affected by the Context of utterance. We hypothesize that the direction of such shift in the mixed condition, if any, is towards a corresponding L1 category. Therefore, we expect the effect of Context on formant frequencies, i.e. the direction of shift, to be moderated by the vowel category (see Section \ref{asymmetry between sounds}, leading to our second research question \textbf{RQ2---Asymmetry}). Thus, the main effect of interest is the interaction term Context*Vowel. This corresponds to the transfer pattern. Our third research question (\bf{RQ3---Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer. If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction. Since we measured FFs at five points in the vowel, Time is expected to affect the FF. However, the precise effect of formant dynamics is expected to depend on the vowel category. Thus, we expect the interaction of Vowel*Time to predict formant frequency. If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.

@@

The sequence of models fitted can be found in the Supplementary materials. The optimal model was: f1 $\sim$ time*vowel + vowel*context*task + time*task + (1 + context|word) + (1 + context + vowel|subject). Table \ref{table_f1} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model. p-vales smaller than 0.05 are starred.

@@

Independent of experimental condition, there is a significant effect of Task: the sentence task (code-switching) leads to higher F1 values in both vowels relative to the cued picture-naming task. This is surprising, as we have no \textit{a priori} reason to expect a uniform difference in vowel quality across tasks. A significant interaction with Time is strongest at the beginning of the vowel, indicating a local effect. A significant Vowel*Context*Task interaction indicates a difference in transfer patterns such that transfer in the F1 (height) dimension is greater in the code-switching (s) task compared to the picture-naming (p) task.\\

@@

\subsection{Vowel backness -- F2}
The optimal model was: f2 $\sim$ vowel*time + vowel*context*task + task*time + (1+context|word) + (1+context+vowel|subject). Table \ref{table_f2} summarizes the model coefficient estimates ($\beta$), standard errors (S.E.), t-values, and p-values for this model. p-values smaller than 0.05 are in starred.

@@

As with F1, we find that Time significantly influences F2 values as expected. Unlike in the F1 dimension, Vowel is a significant predictor of F2, with \nt{2} having significantly lower F2, i.e. being produced as more backed, than \nt{\ae}. There is no significant independent effect of Context, showing that mixed language use does not lead to uniform fronting or backing across vowels. This is not surprising, as we expect movement to be towards the corresponding L1 category, and thus differ across vowels. The negative slope of the estimate indicates that on average, vowels are fronted in the mixed condition. This effect is greater in \nt{\ae}. The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time, suggesting that the underlying acoustic targets are different in the two conditions.

We see a significant effect of Task on F2: vowels in the code-switching (s) task have lower F2, i.e. are more backed than those in the picture-naming (p) task. As with F1, this is an unexpected result. The significant interaction with Time shows that this effect is local and decreases over time. The Vowel*Context*Task interaction is of interest, as it relates to our research question of how the test paradigm affects transfer. Unlike vowel height, we do not find patterns of shift in vowel backness (F2) to differ across paradigms, as indicated by the lack of a significant effect for the fixed effect.

@@

Statistical analyses reveal that the vowel \nt{2} is lowered and fronted, while \nt{\ae} is fronted, in the mixed condition. As discussed in section \ref{asymmetry between sounds}, although our primary research question is about whether productions in switch vs nonswitch conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. Figure \ref{vowels_unilingual_ie_bengali} shows the non-switched IE productions of the participants in relation to average positions of related vowel categories in Bengali. Figure \ref{vowels_e_b} overlays productions in the mixed condition over this, showing F1$\times$F2 vowel plots for IE vowel productions in the nonswitch and mixed conditions, in comparison to the related Bengali categories from the SHRUTI corpus \cite{shruticorpus}. Given that the Bengali corpus data represent average vowel positions, and are not from the same set of speakers as the participants of this study, we refrained from including these in statistical analyses. However, figure \ref{vowels_e_b} provides a heuristic confirmation for our informal hypotheses in section \ref{asymmetry between sounds}) that the IE \nt{\ae} moves towards the corresponding \nt{\ae} category in Bengali, whereas the shift in \nt{2} is towards the related Bengali category \nt{a:}. The following sections discuss the implications of these results.

@@

The aim of this study was to examine short-term phonetic transfer in L2, towards understanding language processing in proficient bilinguals. Two interrelated questions, discussed in section \ref{introduction}, bear on this: (i) what causes transfer in mixed-language speech; and (ii) what causes observed asymmetries between transfer in L1 vs L2.

\paragraph{\textbf{What causes transfer: language mode vs switching}}
In this study, we presented nonswitched English and code-switched Bengali-English stimuli in the same session and found consistent differences between switched and non-switched tokens. This confirms that the local act of switching does trigger transfer, even in a bilingual language mode, paralleling results from \cite{olson2016role, tsui2019impact, piccinini2015voice}. These studies focused on a temporal feature, VOT. Our results indicate that the same happens in vowels. The exact effect of transfer is interesting: since VOT is a one-dimensional measure, observing a change in VOT in a mixed language condition makes it difficult to infer whether a consonant has moved towards a specific L1/L2 counterpart, or reflects a more general system-wide shift in articulatory norms. Vowel quality measures, showing both directionality and magnitude, facilitate this. We find that productions of English vowels do not merely shift towards a Bengali counterpart, but rather towards a set of Bengali-like \textit{norms}: a vowel category that doesn't exist in Bengali nonetheless moves towards a related (perceptually close) Bengali category. This parallels results from \cite{simonet2014phonetic} for Catalan and Spanish (they examined language mode, without code-switching). The study showed that co-activating multiple languages within a discourse causes a transient increase in cross-language phonetic interaction, leading to changes in both L1 and L2. This is interpreted as a follows: a bilingual mode temporarily co-activates both language systems, making both L1 and L2 representations available during processing, whose interactions lead to cross-language transfer. But what does this interaction look like, and how are these abstract representations processed during speech? Examining transfer during switching can provide insights.

The time course of such shifts is telling: we see a difference between the conditions right from the start of articulation, suggesting different acoustic targets. This suggests that such interaction involves the speech planning stage. The pattern of transfer observed in this study, particularly for \nt{2}, also demonstrates that beyond a simple L2-L1 mapping, transfer during online processing makes reference to entities beyond the level of the individual sound, demonstrating system-level interaction: a sound that does not exist in the L1 repertoire nonetheless acquires a different target, and the target is related to the L1 system in a meaningful way. This cannot be reasonably explained as articulatory inertia carried over from uttering L1 sounds. It is also difficult to account for in a purely exemplar framework, as the discourse does not contain any instances of ``L1 \nt{2}" that could affect the L2 category. If representations are combined in creative ways during online speech processing within an utterance, then can we think of intermediate representations not just for individual sounds, but rather for sound systems? While more detailed empirical data is needed to answer such questions, this approach aligns with sociophonetic understandings of bilingual phonetic variation, for example as discussed by \cite{khattab2009phonetic,khattab2013phonetic} based on the use of phonological variables by Arabic-English bilingual children. She visualizes the bilingual phonetic system as ``a rich store of phonological representations", comprising of both language-internal and cross-language variation, from which the bilingual speaker is constantly choosing, guided by a variety of pressures. We propose that vowels are a good feature to examine the plausibility of such systems as cognitive models for language processing: in addition to continuous measures (unlike phonological variables), they also provide two-dimensional data (unlike VOT) that is suited to further test and refine predictions from such a model.

@@

\subsection{Asymmetries in Phonetic Transfer}

Our results indicate differing effects of experimental conditions across the two vowels in both magnitude and direction of shift. Overall, shift is greater in \nt{2}. Recall (figure \ref{vowels_ie}) that the category \nt{\ae} is in a denser part of the IE vowel space than \nt{2}. Specifically: \nt{\ae} is already at the lowermost part of the vowel space. Moreover, large movements upwards in the F1 (vowel height) dimension would risk a conflation of the category with the neighboring \nt{E}. In contrast, \nt{2} (indicated as \nt{@} in the chart) is in the central part of the vowel space and is not flanked by any immediate neighboring categories. Thus, the difference in magnitude of transfer aligns with our hypothesis \textbf{H2---Asymmetry} that the extent of shift in individual categories is constrained by the linguistic principle of contrast preservation. An alternative way to interpret this result is that the initial distance between IE \nt{2} and its target L1 category \nt{a:} was greater than that between the IE and Bengali \nt{\ae} categories, leading to a greater visible shift in the former. However, our results find parallels in previous studies with VOT that have consistently observed greater shifts in the long-lag VOT language (which has a greater acceptable range of VOT) compared to the short-lag language \citep[e.g.][]{antoniou2011inter, bullock2009trying}. This suggests that a general principle of ``room for movement" does play a role in moderating transfer. More broadly, these results collectively suggest that phonetic transfer (i) makes reference to system-level representations that go beyond individual sounds; and (ii) respects linguistic principles. This has implications both for our understanding of phonetic transfer as a phenomenon (e.g. which mental representations and at what level of processing are involved?) and language as a system (e.g. how have transfer phenomena shaped phonological systems in multilingual settings, where languages have co-existed for long periods of time? What are the linguistic implications of increasingly cosmopolitan urban spaces and identities?).

Our results also indicate asymmetry in the direction of shift: the category \nt{2} is lowered and fronted in the mixed language condition, while \nt{\ae} is fronted. On the basis of previous literature indicating that both long-term and short-term shifts in phonological categories of multilingual speakers are targeted with respect to the other language, we hypothesized that participants' IE \nt{\ae} would shift towards the Bengali \nt{\ae}, and \nt{2} towards the Bengali \nt{a:}. As figure \ref{vowels_e_b} indicates, comparing transfer patterns with Bengali categories from \cite{shruticorpus} suggests that this is borne out.
